# Feasibility Assessment

## Executive Summary

*Date:* 2022-11-02

*Summary:* A feasibility assessment to assess candidacy of partner site
location

## Running Feasbility Assessment

### Packages

The following packages will be loaded to conduct the feasibility
assessment:

#### R Packages

You will need the following packages for this tutorial which you can install with `install.packages()`:

- `dplyr`
- `tibble`
- `JuliaConnectoR`

```{r, eval = FALSE, message = FALSE}
library(dplyr)
library(tibble)
library(JuliaConnectoR)
```

To learn more about these packages, see the [Appendix](#appendix).

#### Julia Packages

You will need the following packages for this tutorial:

-   `HealthSampleData`
-   `OMOPCDMDatabaseConnector`
-   `OMOPCDMCohortCreator`
-   `SQLite`
-   `DBInterface`
-   `LibPQ`

To install the Julia packages and activate our analysis environment, we
can run the following code:

```{r Installing Julia Packages, eval = F}
pkg <- juliaImport("Pkg")

pkg$activate("TUTORIAL", shared = TRUE) 
# NOTE: You could specify the path to where your project is or any other path you want; set `shared` to `FALSE` if you do

pkg$add(c("OMOPCDMCohortCreator", "OMOPCDMDatabaseConnector", 
          "HealthSampleData", "SQLite", "DBInterface", "LibPQ"))
```

To learn more about these packages, see the [Appendix](#appendix).

##### Activating Analysis Environment

Now within R, anytime you want to use these installed packages, do the
following:

```{r Activating Julia, eval = F}
library(JuliaConnectoR)

pkg <- juliaImport("Pkg")
pkg$activate("TUTORIAL", shared = TRUE) 
```

### Data

For this tutorial, we will work with the data from Eunomia that is
stored in an SQLite format.

```{r Loading Eunomia, eval = F}
hsd <- juliaImport("HealthSampleData")

eunomia <- hsd$Eunomia()
```

### Defining Connection Details to Eunomia

After you have finished your set up in R, we need to establish a
connection to the Eunomia SQLite database that we will use for the rest
of the tutorial:

```{r Eunomia Connection, eval = F}
slt <- juliaImport("SQLite")

conn <- slt$DB(eunomia)
```

### Defining Connection Details to Other Databases

Here, we need to set-up connection to the OMOP CDM database we will
assess. To do so, we need to define some constants that will be used for
the connection. The following list of constants:

-   dbms - the database management system that is used to host your
    database; common options include (see all options
    [here](http://ohdsi.github.io/DatabaseConnector/reference/createConnectionDetails.html))
    -   `"postgresql"`
    -   `"sqlserver"`
-   server - name of the server; could be localhost, an address like
    123.0.1.5, etc.
-   user - your username to access the server
-   password - the password you use to access the server
-   port - the port where the database is hosted
-   schema - name of the database schema used

Must be defined in this code block:

```{r, eval = FALSE}
dbms <- "Fill in here"
server <- "Fill in here"
user <- "Fill in here"
password <- "Fill in here"
port <- "Fill in here"
schema <- "Fill in here"
```


Additionally, we will need import the Julia packages and activate our
analysis environment:

```{r Other Connections, eval = F}
library(JuliaConnectoR)

pkg <- juliaImport("Pkg")
pkg$activate("TUTORIAL", shared = TRUE)

dbi <- juliaImport("DBInterface")
lq <- juliaImport("LibPQ")
juliaImport("OMOPCDMDatabaseConnector")

```

Now, we can create our connection string using the defined constants
above. The connection string will be in the format of:

```{r, eval = F}
connectionString <- "host=[your host] port=[your port] user=[your user] password=[your password]"
```

With the connection string, we can create the connection to the
database:

```{r, eval = F}
conn <- dbi$connect(lq$Connection, connectionString)
```

#### Example Connection

If any of this was confusing, here is an example of how to fill out the
above connection information, activate our environment, and create a connection:

```{r, eval = F}
dbms <- "postgresql"
server <- "test.data.americus.edu/mimic_omop"
user <- "mimic"
password <- "omoprocks"
port <- 5042
schema <- "mimic.omop"

library(JuliaConnectoR)

pkg <- juliaImport("Pkg")
pkg$activate("TUTORIAL", shared = TRUE)

dbi <- juliaImport("DBInterface")
lq <- juliaImport("LibPQ")
juliaImport("OMOPCDMDatabaseConnector")

connectionString <- paste0("host=", server, " "
                           "port=", port, " "
                           "user=", user, " "
                           "password=", password)

conn <- dbi$connect(lq$Connection, connectionString)

```

### Connecting to the Database and Testing Connection

With Eunomia, the database's schema is simply called "main". We will use
this to generate database connection details that `OMOPCDMCohortCreator`
will use internally. For this step, we will use `OMOPCDMCohortCreator`:

```{r, eval = F}
occ <- juliaImport("OMOPCDMCohortCreator")

occ$GenerateDatabaseDetails(juliaEval(":sqlite"), "main")
```

Finally, we will generate internal representations of each table found
within Eunomia for `OMOPCDMCohortCreator` to use:

```{r, eval = F}
occ$GenerateTables(conn)
```

As a check to make sure everything was correctly installed and works
properly, the following block should work and return a list of all
person ids in this data:

```{r, eval = F}
occ$GetDatabasePersonIDs(conn) 
```

### Characterization Analysis and Queries

#### Background for Analysis

As all the tools are working properly, let's do what is called a
characterization study - a study that *characterizes* a group of
patients with a certain condition (or conditions) across various
attributes like race, age, and combinations thereof.

#### Task: Find All Patients by Condition Concept ID

For this task, we want to load the concept set and get all of the
concept IDs for each condition: bipolar disorder, depression, and
suicidality.

1.  Load condition_concept_ids for a condition

```{r, eval = F}
pathToPhenotype_bipolar <- system.file("phenotypes", "bipolar_concept_set.csv", package ="MentalHealthEquity")
bipolar_concept_ids <- read.csv(pathToPhenotype_bipolar)$CONCEPT_ID

pathToPhenotype_depression <- system.file("phenotypes", "depression_concept_set.csv", package ="MentalHealthEquity")
depression_concept_ids <- read.csv(pathToPhenotype_depression)$CONCEPT_ID

pathToPhenotype_suicidality <- system.file("phenotypes", "suicidality_concept_set.csv", package ="MentalHealthEquity")
suicidality_concept_ids <- read.csv(pathToPhenotype_suicidality)$CONCEPT_ID

```

2.  Filter database patients to those captured by a concept set

```{r, eval = F}
bipolar_patients <- occ$ConditionFilterPersonIDs(bipolar_concept_ids, conn)
bipolar_patients <- bipolar_patients$person_id

depression_patients <- occ$ConditionFilterPersonIDs(depression_concept_ids, conn)
depression_patients <- depression_patients$person_id

suicidality_patients <- occ$ConditionFilterPersonIDs(suicidality_concept_ids, conn)
suicidality_patients <- suicidality_patients$person_id
```

#### Stratified Person Query

Now, we have all the person IDs of database patients captured by each
condition concept set. This next part of the tutorial focuses on
stratifying our database patients by race, gender, and age group. We
will be using the Julia package `OMOPCDMCohortCreator` to characterize
each person, and we will store this information in a tibble to be
exported as a csv.


#### Task 1: Find the race of patients

For this task, we will be using the `GetPatientRace()` function from the `OMOPCDMCohortCreator` package. 

**Bipolar Disorder**

```{r, eval = F}
bipolar_patients_race <- occ$GetPatientRace(bipolar_patients, conn)
```

**Depression**

```{r, eval = F}
depression_patients_race <- occ$GetPatientRace(depression_patients, conn)
```

**Suicidality**

```{r, eval = F}
suicidality_patients_race <- occ$GetPatientRace(suicidality_patients, conn)
```


#### Task 2: Find the gender of patients

For this task, we will be using the `GetPatientGender()` function from the `OMOPCDMCohortCreator` package. 

**Bipolar Disorder**

```{r, eval = F}
bipolar_patients_gender <- occ$GetPatientGender(bipolar_patients, conn)
```

**Depression**

```{r, eval = F}
depression_patients_gender <- occ$GetPatientGender(depression_patients, conn)
```

**Suicidality**

```{r, eval = F}
suicidality_patients_gender <- occ$GetPatientGender(suicidality_patients, conn)
```


#### Task 3: Create Age Groupings of Patients

For this task, for every single person who has a diagnosis of
depression, we need to assign them an age group. For this demo, age
groupings will be made along 55 year intervals when assigned to a person
up to `100` years of age (e.g. `[0, 4], [5, 9], ..., [95,100]`).

```{r, eval = F}
age_groups <- list(
	list(0, 4),
	list(5, 9),
	list(10, 14),
	list(15, 19),
	list(20, 24),
	list(25, 29),
	list(30, 34),
	list(35, 39),
	list(40, 44),
	list(45, 49),
	list(50, 54),
	list(55, 59),
	list(60, 64),
	list(65, 69),
	list(70, 74),
	list(75, 79),
	list(80, 84),
	list(85, 89),
	list(90, 94),
	list(95, 99))

```

**Bipolar Disorder**

```{r, eval = F}
bipolar_patients_age_group <- occ$GetPatientAgeGroup(bipolar_patients, conn, age_groupings = age_groups)
```

**Depression**

```{r, eval = F}
depression_patients_age_group <- occ$GetPatientAgeGroup(depression_patients, conn, age_groupings = age_groups)
```


**Suicidality**

```{r, eval = F}
suicidality_patients_age_group <- occ$GetPatientAgeGroup(suicidality_patients, conn, age_groupings = age_groups)
```

#### Task 4: Characterize Each Person by Race, Gender, and Age Group

With the previous tasks, we now know patients' gender, race, and age group. Using this information, we can combine these features to create a final table showing each patient's person_id, gender, race, and age group per row. To do this combining, we will use the `dplyr` and `tibble` packages:

**Bipolar Disorder**

```{r, eval = F}
library(dplyr)
library(tibble)

bipolar_patients <- tibble(data.frame(person_id=bipolar_patients))
bipolar_patients_race <- tibble(data.frame(bipolar_patients_race))
bipolar_patients_gender <- tibble(data.frame(bipolar_patients_gender))
bipolar_patients_age_group <- tibble(data.frame(bipolar_patients_age_group))
```


**Depression**

```{r, eval = F}
library(dplyr)
library(tibble)

depression_patients <- tibble(data.frame(person_id=depression_patients))
depression_patients_race <- tibble(data.frame(depression_patients_race))
depression_patients_gender <- tibble(data.frame(depression_patients_gender))
depression_patients_age_group <- tibble(data.frame(depression_patients_age_group))
```

**Suicidality**

```{r, eval = F}
library(dplyr)
library(tibble)

suicidality_patients <- tibble(data.frame(person_id=suicidality_patients))
suicidality_patients_race <- tibble(data.frame(suicidality_patients_race))
suicidality_patients_gender <- tibble(data.frame(suicidality_patients_gender))
suicidality_patients_age_group <- tibble(data.frame(suicidality_patients_age_group))
```


#### Task 5: Create Patient Groupings: Join the tables together 

Often with characterization style studies, it is extremely important to
aggregate patient populations. This is to protect the anonymity of
patients with perhaps severely sensitive conditions (e.g. mental
illnesses, sexually transmitted diseases, etc.) from possible
repercussions from accidental disclosure of this patient information.

**Bipolar Disorder**

```{r, eval = F}
bipolar_final_df <- full_join(bipolar_patients, bipolar_patients_race, by = c("person_id" = "person_id")) %>% 
  full_join(bipolar_patients_age_group, by = c("person_id" = "person_id")) %>%
  full_join(bipolar_patients_gender, by = c("person_id" = "person_id")) %>%
  select(-person_id) %>% 
  count(race_concept_id, age_group, gender_concept_id) %>%
  rename(count = "n") %>%
  filter(count > 10)
```


**Depression**

```{r, eval = F}
depression_final_df <- full_join(depression_patients, depression_patients_race, by = c("person_id" = "person_id")) %>% 
  full_join(depression_patients_age_group, by = c("person_id" = "person_id")) %>%
  full_join(depression_patients_gender, by = c("person_id" = "person_id")) %>%
  select(-person_id) %>% 
  count(race_concept_id, age_group, gender_concept_id) %>%
  rename(count = "n") %>%
  filter(count > 10)
```

**Suicidality**

```{r, eval = F}
suicidality_final_df <- full_join(suicidality_patients, suicidality_patients_race, by = c("person_id" = "person_id")) %>% 
  full_join(suicidality_patients_age_group, by = c("person_id" = "person_id")) %>%
  full_join(suicidality_patients_gender, by = c("person_id" = "person_id")) %>%
  select(-person_id) %>% 
  count(race_concept_id, age_group, gender_concept_id) %>%
  rename(count = "n") %>%
  filter(count > 10)
```


### Conclusion

This concludes the tutorial for our characterization study in Eunomia. This mini characterization study that we just conducted on this dataset opens up a whole new avenue for a researcher to pursue. For example, we could now calculate prevalence rates across different patient characteristics or compare and contrast multiple conditions at once. It should also be apparent that the API is set up in a very particular way: it is functional meaning that each function does one thing only. This gives a lot of flexibility to a user to build together a study incrementally using `OMOPCDMCohortCreator.`


### Appendix

#### Packages Used in Analysis:

##### R Packages Used:

- `dplyr` - grammar of data manipulation
- `tibble` - modern reimagining of the data.frame
- `JuliaConnectoR` - a functionally oriented interface for calling Julia from R


##### Julia Packages Used:

-   `HealthSampleData` - Sample health data for a variety of health formats and use cases
-   `OMOPCDMDatabaseConnector` - Connect to databases utilizing the OMOP CDM 
-   `OMOPCDMCohortCreator` - Create cohorts from databases utilizing the OMOP CDM
-   `SQLite` - Julia interface to the SQLite library
-   `DBInterface` - Julia interface for common database operations and methods
-   `LibPQ` - Julia interface to the PostgreSQL library
