# Testing Instructions

## Executive Summary 

**Author:** Jacob S. Zelko

**Date:** 2022-11-18

**Summary:** A small instructional that ensures the environment set-up ran correctly

## Environment Set-Up

### Configuring R Environment

Open RStudio and navigate to the folder `MentalHealthEquity/feasibility`.
Within that directory, do the following steps within the RStudio console: 

1. Load an RStudio project within the `MentalHealthEquity/feasibility` folder 
2. Load the `renv` package: `library(renv)` 
3. Activate the `renv` environment: `renv::activate()`
4. Restore the needed project packages by running: `renv::restore()` and accepting 

The following code block shows how this would run:

```{r, eval = FALSE}
library(renv)

renv::activate()
renv::restore()
```

### Executing Tests

From within the project, run the following cell:

```{r, eval = FALSE}
source("tests/test_environment.R")
```

You should see some output from this cell and announcements about tests running.
If everything ran correctly on your machine, then you should see messages about the tests passing as well as celebratory emojis!
You will then be able to continue to the next step of the assessment!

