---
title: "Characterize a Given OMOP CDM Database"
output: pdf_document
---
# Initial Set-Up

## Activating Project Environment

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = F, results = 'hold')
options(scipen=999) 
devtools::load_all()
library(MentalHealthEquity)
```

## Packages Used in Analysis

`r library(DatabaseConnector)`

`DatabaseConnector`

Package Description:

This package will be used to create connections to the database and execute queries to obtain characterizations about the cohorts in the database. 

# Introduction

## Background

The intent of this report is to investigate the characterization statement:

> **Characterization Statement 1:** Characterize the individuals being seen for mental health care services (related to depression, bipolar disorder, and suicidality) at least one time – including hospitalization events.

This characterization statement is founded on the central research topic for this study:

> **Research Topic:** Based on [CLAIMS], we see X% of all persons with at least one claim indicating [DEPRESSION/BIPOLAR DISORDER/SUICIDALITY] are not seen again.

By which the phrases "…all persons…" refer to those seen by patience care provider, etc. and "…are not seen again." implies lack of adherence to care.


# Data Analysis Preparation

## Creating Initial Connection

Defining connection details for connecting to a given database:

```{r Establishing Connection}
connectionDetails <- createConnectionDetails(
  dbms = "postgresql",
  server = "data.hdap.gatech.edu/synpuf_v5",
  user = keyring::key_list("synPUF")[1,2],
  password = keyring::key_get("synPUF", keyring::key_list("synPUF")[1,2]),
  port = 5434,
  pathToDriver = "../utils"
)
```


## Creating Initial OMOP Tables

In this case, the schema follows the [OMOP CDM v5](https://ohdsi.github.io/CommonDataModel/index.html) schema:

```
```

For this analysis, we will work with the following tables from the schema:

```
```

<!--TODO: Add description on what these tables are from the Book of OHDSI-->
Per the [Book of OHDSI], here are the break downs for these tables:

- `PERSON` - 
- `LOCATION` - 
- `OBSERVATION_PERIOD` - 
- `CONDITION_OCCURRENCE` -



## General Characterization of Data

```{r Joined SQL for Demographics}
sql <- "
  SELECT @table.@concept_name, count(*)
  FROM person 
      LEFT JOIN (select LOCATION_ID, STATE FROM location) AS location ON 
        person.location_id = location.location_id
      LEFT JOIN (select * FROM concept WHERE DOMAIN_ID = 'Race' 
              AND STANDARD_CONCEPT = 'S') AS race ON
        person.race_concept_id = race.concept_id
      LEFT JOIN (select * FROM concept WHERE DOMAIN_ID = 'Ethnicity' 
              AND STANDARD_CONCEPT = 'S') AS ethnicity ON
        person.ethnicity_concept_id = ethnicity.concept_id
      LEFT JOIN (select * FROM concept WHERE DOMAIN_ID = 'Gender' 
              AND STANDARD_CONCEPT = 'S') AS gender ON
        person.gender_concept_id = gender.concept_id
  GROUP BY @table.@concept_name
  ORDER BY count(*) DESC
"
```

# Cohort Populations

## General Cohort

### Overall Trends

```{r Overall Admit Dates}
conn <- connect(connectionDetails)
date_sql <- "
  SELECT  MIN(visit.visit_end_date) as first_visit,
          MAX(visit.visit_end_date) as last_visit
  FROM visit_occurrence visit
"
dates <- renderTranslateQuerySql(conn, date_sql)
disconnect(conn)
dates
```

The dates in the dataset range from `r dates$FIRST_VISIT` and `r dates$LAST_VISIT`

#### Patient Spread

This calculates the number of patients in the dataset:

```{r Overall Count}
conn <- connect(connectionDetails)
count <- querySql(conn, sql = "SELECT COUNT(DISTINCT PERSON_ID) FROM PERSON")
disconnect(conn)
```

In this dataset, there are `r count` patients.
To further examine this data, we can break them down across the following axes:

- State
- Age
- Race
- Gender
- Care setting


##### State Breakdown

```{r Overall Location}
conn <- connect(connectionDetails)
state <- renderTranslateQuerySql(conn, sql, table = "location", concept_name = "state")
disconnect(conn)

```

The top states are: `r head(state, n = 10)$STATE`

`r knitr::kable(head(state, n = 10))`


##### Age Breakdown

```{r Overall Age of Admit}
conn <- connect(connectionDetails)
age_sql <- "
  SELECT  MIN(YEAR(visit.visit_end_date) - person.year_of_birth) as min_age, 
          MAX(YEAR(visit.visit_end_date) - person.year_of_birth) as max_age
  FROM person
    JOIN visit_occurrence visit ON
      person.person_id = visit.person_id
"
age <- renderTranslateQuerySql(conn, age_sql)
disconnect(conn)
age
```

The age of the patient at the time of visit ranged from `r dates$MIN_AGE` years of age to `r dates$MAX_AGE` years of age. 


##### Racial Breakdown

```{r Overall Race}
conn <- connect(connectionDetails)
race <- renderTranslateQuerySql(conn, sql, table = "race", concept_name= "concept_name")
disconnect(conn)
race
```

In the dataset, there are `r mapply(paste, race$COUNT, race$CONCEPT_NAME, "persons")`
`r knitr::kable(race)`


##### Gender Breakdown

```{r Overall Gender}
conn <- connect(connectionDetails)
gender <- renderTranslateQuerySql(conn, sql, table = "gender", concept_name = "concept_name")
disconnect(conn)
gender
```

In the dataset, there are `r mapply(paste, gender$COUNT, gender$CONCEPT_NAME, "persons")`
`r knitr::kable(gender)`


##### Care Site Breakdown

```{r Overall Care Site}
conn <- connect(connectionDetails)
caresite_sql <- "
  SELECT concept.concept_name, count(*)
  FROM visit_occurrence visit 
    JOIN (SELECT CONCEPT_NAME , CONCEPT_ID
          FROM concept 
          WHERE DOMAIN_ID = 'Visit' 
              AND STANDARD_CONCEPT = 'S') concept ON
      visit.visit_concept_id = concept.concept_id
  GROUP BY concept.concept_name
  ORDER BY count(*) DESC
"
care_site <- renderTranslateQuerySql(conn, caresite_sql)
disconnect(conn)
care_site
```

In the dataset, the care site setting `r mapply(paste, care_site$CONCEPT_NAME, "had", care_site$COUNT, "visits")`
