---
title: "Characterize a Given OMOP CDM Database"
output: pdf_document
---
# Initial Set-Up

## Activating Project Environment

```{r setup, include = F}
knitr::opts_chunk$set(echo = F, message = F, results = 'hold', cache = T, fig.align = 'center', fig.height = 3)
options(scipen=999) 
devtools::load_all()
library(MentalHealthEquity)
library(dplyr)
library(ggplot2)
library(forcats) #refactoring
```

## Packages Used in Analysis

```{r Libraries and Packages}
library(DatabaseConnector)
```

`DatabaseConnector`

Package Description:

This package will be used to create connections to the database and execute queries to obtain characterizations about the cohorts in the database. 

# Analysis Introduction

## Background on Analysis

The intent of this report is to investigate the characterization statement:

> **Characterization Statement 1:** Characterize the individuals being seen for mental health care services (related to depression, bipolar disorder, and suicidality) at least one time – including hospitalization events.

This characterization statement is founded on the central research topic for this study:

> **Research Topic:** Based on [CLAIMS], we see X% of all persons with at least one claim indicating [DEPRESSION/BIPOLAR DISORDER/SUICIDALITY] are not seen again.

By which the phrases "…all persons…" refer to those seen by patience care provider, etc. and "…are not seen again." implies lack of adherence to care.


# Data Analysis Preparation

## Creating Initial Connection

Defining connection details for connecting to a given database:

```{r Establishing Connection, include = T}
connectionDetails <- createConnectionDetails(
  dbms = "postgresql",
  server = "data.hdap.gatech.edu/synpuf_v5",
  user = keyring::key_list("synPUF")[1,2],
  password = keyring::key_get("synPUF", keyring::key_list("synPUF")[1,2]),
  port = 5434,
  pathToDriver = "../utils"
)
```


## Creating Initial OMOP Tables

In this case, the schema follows the [OMOP CDM v5](https://ohdsi.github.io/CommonDataModel/index.html) schema.

For this analysis, we will work with the following tables from the schema:

- `PERSON` 
- `LOCATION` 
- `OBSERVATION_PERIOD` 
- `CONDITION_OCCURRENCE` 

## Defining Conditions
Using the concept_ids for bipolar, depression, and suicidality, we can grab the concept set related to the each condition.

```{R Defining Conditions, include = T}
bipolar_id <- 436665
bipolar_df <- get_atlas_concept_set(get_atlas_concept(bipolar_id))
bipolar_concept <- paste0('(', paste(bipolar_df$CONCEPT_ID, collapse = ", "), ')')

depression_id <- 440383
depression_df <- get_atlas_concept_set(get_atlas_concept(depression_id))
depression_concept <- paste0('(', paste(depression_df$CONCEPT_ID, collapse = ", "), ')')

suicidality_id <- 4273391
suicidality_df <- get_atlas_concept_set(get_atlas_concept(suicidality_id))
suicidality_concept <- paste0('(', paste(suicidality_df$CONCEPT_ID, collapse = ", "), ')')
```


## General Characterization of Data

```{r Overall Count}
conn <- connect(connectionDetails)
count <- querySql(conn, sql = "SELECT COUNT(DISTINCT PERSON_ID) FROM PERSON")
disconnect(conn)
```

```{r Overall Admit Dates}
conn <- connect(connectionDetails)
date_sql <- "
  SELECT  MIN(visit.visit_end_date) as first_visit,
          MAX(visit.visit_end_date) as last_visit
  FROM visit_occurrence visit
"
dates <- renderTranslateQuerySql(conn, date_sql)
disconnect(conn)
```


In this dataset, there are `r count` patients. The dates in the dataset range from `r dates$FIRST_VISIT` and `r dates$LAST_VISIT`.
To further examine this data, we can break them down across the following axes:

- State
- Age
- Race
- Gender
- Care setting

#### Workflow for Location/State, Race, Ethnicity, and Gender

To find the the demographics of the individuals in the data set, I created a parameterized SQL containing three parameters: the table/domain, concept name (demographic is default concept name, but location is diff), and a bool var for specific demographic or entire table

1. Select the `person` table

2. Join the `person` table with the `location` table joining on `location_id`

3. Join with `concept` table using the different `concept_id` to find the names of the concept ids in the person table

4. Group by demographic (location, race, ethnicity, gender, etc)

5. Return the counts of the patients by demographics



```{r Joined SQL for Demographics}
sql <- "
  SELECT {@x} ? {@table.@concept_name, count(DISTINCT PERSON_ID)} : {*}
  FROM person 
      LEFT JOIN (SELECT LOCATION_ID, STATE FROM location) AS location ON 
        person.location_id = location.location_id
      LEFT JOIN (SELECT CONCEPT_ID, 
                    CONCEPT_NAME as RACE_CONCEPT_NAME  
                  FROM concept WHERE DOMAIN_ID = 'Race' 
                  AND STANDARD_CONCEPT = 'S') AS race ON
        person.race_concept_id = race.concept_id
      LEFT JOIN ( SELECT CONCEPT_ID, 
                    CONCEPT_NAME as ETHNICITY_CONCEPT_NAME 
                  FROM concept WHERE DOMAIN_ID = 'Ethnicity' 
                    AND STANDARD_CONCEPT = 'S') AS ethnicity ON
        person.ethnicity_concept_id = ethnicity.concept_id
      LEFT JOIN (SELECT CONCEPT_ID, 
                    CONCEPT_NAME as GENDER_CONCEPT_NAME 
                  FROM concept WHERE DOMAIN_ID = 'Gender' 
                  AND STANDARD_CONCEPT = 'S') AS gender ON
        person.gender_concept_id = gender.concept_id
  {@x}? {GROUP BY @table.@concept_name
        ORDER BY COUNT DESC} :
        {LIMIT 10}
"
```


#### State Breakdown

```{r Overall Location}
conn <- connect(connectionDetails)
state <- renderTranslateQuerySql(conn, sql, table = "location", 
                                 concept_name = "state", x = T)
disconnect(conn)

state %>%
  mutate(STATE = fct_reorder(STATE, COUNT, .desc = F)) %>%
  top_n(20) %>%
  ggplot(aes(x = STATE, y = COUNT)) +
  geom_bar(stat = "identity", fill = "steelblue")+
  labs(title = "Overall State Distribution")+
  theme_minimal()
```

The top states are: `r head(state, n = 10)$STATE`

`r knitr::kable(head(state, n = 10))`



#### Racial Breakdown

```{r Overall Race}
conn <- connect(connectionDetails)
race <- renderTranslateQuerySql(conn, sql, table = "race", 
                                concept_name= "RACE_CONCEPT_NAME", x = T)
disconnect(conn)

race %>%
    mutate(RACE_CONCEPT_NAME = fct_reorder(RACE_CONCEPT_NAME, COUNT, .desc = T)) %>%
    ggplot(aes(x = RACE_CONCEPT_NAME, y = COUNT)) +
    geom_bar(stat = "identity", fill = "steelblue")+
    theme_minimal() +
    labs(title = "Overall Race Distribution")+
    theme(axis.title.x = element_blank())
```

In the dataset, there are `r mapply(paste, race$COUNT, race$RACE_CONCEPT_NAME, "persons")`
`r knitr::kable(race)`


#### Gender Breakdown

```{r Overall Gender}
conn <- connect(connectionDetails)
gender <- renderTranslateQuerySql(conn, sql, table = "gender", 
                                  concept_name = "GENDER_CONCEPT_NAME", x = T)
disconnect(conn)

gender %>%
    mutate(GENDER_CONCEPT_NAME = fct_reorder(GENDER_CONCEPT_NAME, COUNT, .desc = T)) %>%
    ggplot(aes(x = GENDER_CONCEPT_NAME, y = COUNT)) +
    geom_bar(stat = "identity", fill = "steelblue")+
    theme_minimal() +
    labs(title = "Overall Gender Distribution")+
    theme(axis.title.x = element_blank())
```

In the dataset, there are `r mapply(paste, gender$COUNT, gender$GENDER_CONCEPT_NAME, "persons")`
`r knitr::kable(gender)`

#### Age Breakdown

To find the range of ages of the dataset:

1. Select the `visit_occurrence` table

2. Join on the `person` table using `person_id`

3. Subtract the `visit_end_date` from the `year_of_birth` 

4. Report the minimum age and maximum age of the person


```{r Overall Age of Admit}
conn <- connect(connectionDetails)
age_sql <- "
  SELECT  MIN(YEAR(visit.visit_end_date) - person.year_of_birth) as min_age, 
          MAX(YEAR(visit.visit_end_date) - person.year_of_birth) as max_age
  FROM person
    JOIN visit_occurrence visit ON
      person.person_id = visit.person_id
"
age <- renderTranslateQuerySql(conn, age_sql)
disconnect(conn)
```

The age of the patient at the time of visit ranged from `r age$MIN_AGE` years of age to `r age$MAX_AGE` years of age. 


#### Care Site Breakdown

To understand more about the care site breakdowns of the dataset: 

1. Select the `visit_occurrence` table

2. Join with the `concept` table to get the concept names using `concept_id`

3. Group by the care site settings

4. Return the counts distinct persons for the care site  


```{r Overall Care Site}
conn <- connect(connectionDetails)
caresite_sql <- "
  SELECT concept.concept_name, count(*) VISIT_COUNT, 
    count(DISTINCT visit.person_id) PERSON_COUNT
  FROM visit_occurrence visit 
    JOIN (SELECT CONCEPT_NAME , CONCEPT_ID
          FROM concept 
          WHERE DOMAIN_ID = 'Visit' 
              AND STANDARD_CONCEPT = 'S') concept ON
      visit.visit_concept_id = concept.concept_id
  GROUP BY concept.concept_name
  ORDER BY count(*) DESC
"
care_site <- renderTranslateQuerySql(conn, caresite_sql)
disconnect(conn)


# care_site %>%
#   ggplot(aes(x = CONCEPT_NAME, y = VISIT_COUNT)) +
#   geom_bar(stat = "identity", fill = "steelblue")+
#   labs(title = "Overall Care Site Distribution")
```

In the dataset, there are `r care_site$PERSON_COUNT` unique patients with an indicated care site. The care site setting `r mapply(paste, care_site$CONCEPT_NAME, "had", care_site$VISIT_COUNT, "visits")` 
`r knitr::kable(care_site)`


## Inpatient Cohort

#### Patient Spread

To find the characterization of the inpatient persons in the dataset: 

1. Select the `visit_occurrence` table 

2. Join on the `person` table using `person_id`

3. Select the different demographics and join them on `concept` using `concept_id` to find the names of the race, ethnicity, and gender concept IDs. 


```{r Inpatient Total}
conn <- connect(connectionDetails)
inpatient_sql <- "
  SELECT @query
  FROM (SELECT *
        FROM visit_occurrence visit
        JOIN
          ( SELECT CONCEPT_NAME , CONCEPT_ID
            FROM concept
            WHERE DOMAIN_ID = 'Visit'
            AND STANDARD_CONCEPT = 'S') concept
        ON visit.VISIT_CONCEPT_ID = concept.CONCEPT_ID
        WHERE LOWER(concept.CONCEPT_NAME) LIKE '%@a%'
        ) VISIT_INFO
  JOIN
    ( SELECT *
      FROM person
          LEFT JOIN (SELECT LOCATION_ID, STATE FROM location) AS location ON
            person.location_id = location.location_id
          LEFT JOIN (SELECT CONCEPT_ID,
                        CONCEPT_NAME as RACE_CONCEPT_NAME
                      FROM concept WHERE DOMAIN_ID = 'Race'
                      AND STANDARD_CONCEPT = 'S') AS race ON
            person.race_concept_id = race.concept_id
          LEFT JOIN ( SELECT CONCEPT_ID,
                        CONCEPT_NAME as ETHNICITY_CONCEPT_NAME
                      FROM concept WHERE DOMAIN_ID = 'Ethnicity'
                        AND STANDARD_CONCEPT = 'S') AS ethnicity ON
            person.ethnicity_concept_id = ethnicity.concept_id
          LEFT JOIN (SELECT CONCEPT_ID,
                        CONCEPT_NAME as GENDER_CONCEPT_NAME
                      FROM concept WHERE DOMAIN_ID = 'Gender'
                      AND STANDARD_CONCEPT = 'S') AS gender ON
            person.gender_concept_id = gender.concept_id) DEMO_INFO
  USING(PERSON_ID)
  {@x} ? {GROUP BY DEMO_INFO.@concept_name
          ORDER BY COUNT DESC}
"
inpatient <- renderTranslateQuerySql(conn, 
                                     inpatient_sql, 
                                     x = F, a = "inpatient", 
                                     query = "COUNT(*) VISIT_COUNT, 
                                     COUNT(DISTINCT PERSON_ID) PERSON_COUNT
                                     ")
disconnect(conn)
```

In the inpatient cohort, there are a total of `r inpatient$PERSON_COUNT` distinct patients, and there are a total of `r inpatient$VISIT_COUNT` entries. Compared to the original unique person count in the overall dataset, which is `r count`, `r count - inpatient$PERSON_COUNT` are not accounted for in the inpatient count. 

To further examine this data, we can break them down across the following axes:

- State
- Age
- Race
- Gender
- Care setting

#### State Breakdown
```{r Inpatient State}
conn <- connect(connectionDetails)
inpatient_state <- renderTranslateQuerySql(conn, inpatient_sql, x = T, a = "inpatient",
                        query = "DEMO_INFO.state, COUNT(DISTINCT PERSON_ID)",
                        concept_name = "state")
disconnect(conn)

inpatient_state %>%
  mutate(STATE = fct_reorder(STATE, COUNT, .desc = F)) %>%
  top_n(20) %>%
  ggplot(aes(x = STATE, y = COUNT)) +
  geom_bar(stat = "identity", fill = "steelblue")+
  labs(title = "Inpatient State Distribution")+
  theme_minimal()
```

In the inpatient cohort, the top 5 states by person count are 
`r knitr::kable(head(inpatient_state, n = 10))`

#### Age Breakdown

For age breakdowns, inpatient counts were bucketed based on their age. The age brackets were as follows:
- 0-9
- 10-19
- 20-29
- 30-39
- 40-49
- 50-59
- 60-69
- 70-79
- 80-89
- $\ge$ 90

To find the age breakdown of the dataset:

1. Select the `visit_occurrence` table

2. Join on the `person` table using `person_id`

3. Subtract the `visit_end_date` from the `year_of_birth` 


```{r Inpatient Age} 
conn <- connect(connectionDetails)
inpatient_age <- renderTranslateQuerySql(conn, inpatient_sql,
                                   x = F, a = "inpatient",
                                   query = "
                SUM(CASE WHEN YEAR(VISIT_INFO.visit_end_date) - 
                  DEMO_INFO.year_of_birth <= 9 THEN 1 END) AS age_0_9,
                SUM(CASE WHEN YEAR(VISIT_INFO.visit_end_date) -                              
                  DEMO_INFO.year_of_birth BETWEEN 10 AND 19 THEN 1 END) AS age_10_to_19,
                SUM(CASE WHEN YEAR(VISIT_INFO.visit_end_date) -                              
                  DEMO_INFO.year_of_birth BETWEEN 20 AND 29 THEN 1 END) AS age_20_to_29,
                SUM(CASE WHEN YEAR(VISIT_INFO.visit_end_date) -                              
                  DEMO_INFO.year_of_birth BETWEEN 30 AND 39 THEN 1 END) AS age_30_to_39,
                SUM(CASE WHEN YEAR(VISIT_INFO.visit_end_date) -                              
                  DEMO_INFO.year_of_birth BETWEEN 40 AND 49 THEN 1 END) AS age_40_to_49,
                SUM(CASE WHEN YEAR(VISIT_INFO.visit_end_date) -                              
                  DEMO_INFO.year_of_birth BETWEEN 50 AND 59 THEN 1 END) AS age_50_to_59,
                SUM(CASE WHEN YEAR(VISIT_INFO.visit_end_date) -                              
                  DEMO_INFO.year_of_birth BETWEEN 60 AND 69 THEN 1 END) AS age_60_to_69,
                SUM(CASE WHEN YEAR(VISIT_INFO.visit_end_date) -                              
                  DEMO_INFO.year_of_birth BETWEEN 70 AND 79 THEN 1 END) AS age_70_to_79,
                SUM(CASE WHEN YEAR(VISIT_INFO.visit_end_date) -                              
                  DEMO_INFO.year_of_birth BETWEEN 80 AND 89 THEN 1 END) AS age_80_to_89,
                SUM(CASE WHEN YEAR(VISIT_INFO.visit_end_date) -                              
                  DEMO_INFO.year_of_birth >= 90 THEN 1 ELSE 0 END) AS age_90_and_more")
disconnect(conn)
```

In the inpatient dataset, it takes in the visit encounters, and the ages for the individuals were found according to the time of the inpatient visit. 

This table shows the number of occurences of the ages in the data set. 

`r knitr::kable(t(inpatient_age), col.names = "COUNT")`





#### Racial Breakdown
```{r Inpatient Race}
conn <- connect(connectionDetails)
inpatient_race <- renderTranslateQuerySql(conn, inpatient_sql, 
                                          x = T, 
                                          a = "inpatient", 
                                          concept_name = "RACE_CONCEPT_NAME", 
                                          query = "
                                          DEMO_INFO.RACE_CONCEPT_NAME, 
                                          COUNT(DISTINCT PERSON_ID)")
disconnect(conn)



inpatient_race %>%
    mutate(RACE_CONCEPT_NAME = fct_reorder(RACE_CONCEPT_NAME, COUNT, .desc = T)) %>%
    ggplot(aes(x = RACE_CONCEPT_NAME, y = COUNT)) +
    geom_bar(stat = "identity", fill = "steelblue")+
    theme_minimal() +
    labs(title = "Inpatient Race Distribution")+
    theme(axis.title.x = element_blank())
```

In the inpatient cohort, there are `r mapply(paste, inpatient_race$COUNT, inpatient_race$RACE_CONCEPT_NAME, "persons")`

`r knitr::kable(inpatient_race)`


#### Gender Breakdown

```{r Inpatient Gender}
conn <- connect(connectionDetails)
inpatient_gender <- renderTranslateQuerySql(conn, inpatient_sql, 
                        x = T, a = "inpatient", 
                        concept_name = "GENDER_CONCEPT_NAME", 
                        query = "
                        DEMO_INFO.GENDER_CONCEPT_NAME, 
                        COUNT(DISTINCT PERSON_ID)
                        ")
disconnect(conn)


inpatient_gender %>%
    mutate(GENDER_CONCEPT_NAME = fct_reorder(GENDER_CONCEPT_NAME, COUNT, .desc = T)) %>%
    ggplot(aes(x = GENDER_CONCEPT_NAME, y = COUNT)) +
    geom_bar(stat = "identity", fill = "steelblue")+
    theme_minimal() +
    labs(title = "Inpatient Gender Distribution")+
    theme(axis.title.x = element_blank())
```
In the inpatient cohort, there are 
`r mapply(paste, inpatient_gender$COUNT, inpatient_gender$GENDER_CONCEPT_NAME, "persons")`
`r knitr::kable(inpatient_gender)`

## Condition Breakdown
To find the characterize the dataset by condition, I created a parameterized SQL that filters by the condition using the concept set, and it takes four parameters

- A list of concept ids for the condition

- X: a boolean value for an unspecified query (default True means return table)

- If X is FALSE, then query must be specified 

- Y: A boolean value for a grouping the dataset

- If Y is TRUE, then a group by query must be specified 


### General Trends for Bipolar Disorder

```{r Bipolar Cohort}
bipolar_id <- 436665
bipolar_df <- get_atlas_concept_set(get_atlas_concept(bipolar_id))
bipolar_concept <- paste0('(', paste(bipolar_df$CONCEPT_ID, collapse = ", "), ')')
condition_sql <- "
  SELECT {@x} ?
      { person.PERSON_ID, 
        person.YEAR_OF_BIRTH, 
        person.MONTH_OF_BIRTH, 
        person.DAY_OF_BIRTH, 
        location.STATE, 
        race.RACE_CONCEPT_NAME, 
        ethnicity.ETHNICITY_CONCEPT_NAME, 
        gender.GENDER_CONCEPT_NAME, 
        condition.CONDITION_OCCURRENCE_ID, 
        condition.CONDITION_CONCEPT_ID, 
        condition.CONDITION_CONCEPT_NAME,
        condition.CONDITION_START_DATE, 
        condition.CONDITION_END_DATE, 
        condition.VISIT_OCCURRENCE_ID, 
        visit_info.VISIT_CONCEPT_NAME} :
      {@query}
  FROM person 
      LEFT JOIN (SELECT LOCATION_ID, STATE FROM location) AS location ON
        person.location_id = location.location_id
      LEFT JOIN (SELECT CONCEPT_ID,
                    CONCEPT_NAME as RACE_CONCEPT_NAME
                  FROM concept WHERE DOMAIN_ID = 'Race'
                  AND STANDARD_CONCEPT = 'S') AS race ON
        person.race_concept_id = race.concept_id
      LEFT JOIN ( SELECT CONCEPT_ID,
                    CONCEPT_NAME as ETHNICITY_CONCEPT_NAME
                  FROM concept WHERE DOMAIN_ID = 'Ethnicity'
                    AND STANDARD_CONCEPT = 'S') AS ethnicity ON
        person.ethnicity_concept_id = ethnicity.concept_id
      LEFT JOIN ( SELECT CONCEPT_ID,
                  CONCEPT_NAME as GENDER_CONCEPT_NAME
                  FROM concept WHERE DOMAIN_ID = 'Gender'
                      AND STANDARD_CONCEPT = 'S') AS gender ON
        person.gender_concept_id = gender.concept_id
      LEFT JOIN ( SELECT *
                  FROM CONDITION_OCCURRENCE 
                  LEFT JOIN  
                    ( SELECT CONCEPT_ID,
                      CONCEPT_NAME as CONDITION_CONCEPT_NAME
                      FROM concept WHERE DOMAIN_ID = 'Condition'
                          AND STANDARD_CONCEPT = 'S') concept ON 
                  CONDITION_OCCURRENCE.CONDITION_CONCEPT_ID = concept.CONCEPT_ID) condition ON
        person.PERSON_ID = condition.PERSON_ID
      LEFT JOIN ( SELECT *
                  FROM VISIT_OCCURRENCE visit
                  JOIN
                        ( SELECT concept.CONCEPT_NAME AS VISIT_CONCEPT_NAME, 
                              concept.CONCEPT_ID
                          FROM concept
                          WHERE DOMAIN_ID = 'Visit'
                          AND STANDARD_CONCEPT = 'S') concept
                  ON visit.VISIT_CONCEPT_ID = concept.CONCEPT_ID) as visit_info ON 
                  person.PERSON_ID = visit_info.PERSON_ID AND
                  condition.VISIT_OCCURRENCE_ID = visit_info.VISIT_OCCURRENCE_ID
  WHERE condition.CONDITION_CONCEPT_ID in @ids
  {@y}? {@query2} 
"
conn <- connect(connectionDetails)
bipolar_counts <- renderTranslateQuerySql(conn, condition_sql, 
                        ids = bipolar_concept, x = F, y = F, 
                        query = "COUNT(*) AS RECORDS, COUNT(DISTINCT person.PERSON_ID) AS COUNT")
disconnect(conn)
```

#### Patient Spread

This calculates the number of inpatient patients in the dataset:

In this dataset, there are a total of `r bipolar_counts$RECORDS` records of bipolar disorder in the dataset, with `r bipolar_counts$COUNT` patients

To further examine this data, we can break them down across the following axes:

- Condition
- State
- Age
- Race
- Gender
- Care setting


#### Condition Breakdown

For the bipolar characterization:

1. Select the `person` table

2. Join `condition_occurrence` using `person_id`

3. Using the concept set definition for bipolar, filter the persons table for the persons with a bipolar diagnosis

4. Join the `visit_occurrence` table using `person_id` 

5. Select the demographic of interest (location, race, ethnicity, or gender)

6. Group by the demographic 

7. Return the person counts with bipolar for each group


```{r Bipolar Conditions}
conn <- connect(connectionDetails)
bipolar_conditions <- renderTranslateQuerySql(conn, condition_sql, ids = bipolar_concept, x = F,
                                              query = " condition.CONDITION_CONCEPT_NAME, COUNT(*) RECORD_COUNT, COUNT(DISTINCT person.PERSON_ID) PERSON_COUNT", y = T, query2 = "GROUP BY condition.CONDITION_CONCEPT_NAME ORDER BY COUNT(*) DESC")
disconnect(conn)

```

Here is the breakdown of the top bipolar diagnoses: `r knitr::kable(head(bipolar_conditions))`


#### State Breakdown
```{r Bipolar State}
conn <- connect(connectionDetails)
bipolar_state <- renderTranslateQuerySql(conn, condition_sql, ids = bipolar_concept, x = F, 
                        query = "location.STATE, COUNT(DISTINCT person.PERSON_ID)",
                        y = T, query2 = "GROUP BY location.STATE ORDER BY COUNT DESC")
disconnect(conn)

bipolar_state %>%
  mutate(STATE = fct_reorder(STATE, COUNT, .desc = F)) %>%
  top_n(20) %>%
  ggplot(aes(x = STATE, y = COUNT)) +
  geom_bar(stat = "identity", fill = "steelblue")+
  labs(title = "Bipolar State Distribution")+
  theme_minimal()
```

The top ten states by person count with bipolar disorder are
`r knitr::kable(head(bipolar_state, n = 10))`

#### Age Breakdown

For age breakdowns, inpatient counts were bucketed based on their age. The age brackets were as follows:
- 0-9
- 10-19
- 20-29
- 30-39
- 40-49
- 50-59
- 60-69
- 70-79
- 80-89
- $\ge$ 90

```{r Bipolar Age}
conn <- connect(connectionDetails)
bipolar_age <- renderTranslateQuerySql(conn, 
                                       condition_sql, 
                                       ids = bipolar_concept, x = F,
                                       query = "
                MIN(YEAR(condition.CONDITION_START_DATE) -                                   
                  person.year_of_birth) as min_age, 
                MAX(YEAR(condition.CONDITION_START_DATE) - person.year_of_birth) 
                  as max_age, 
                SUM(CASE WHEN YEAR(condition.CONDITION_START_DATE) - 
                  person.year_of_birth <= 9 THEN 1 END) AS age_0_9,
                SUM(CASE WHEN YEAR(condition.CONDITION_START_DATE) -                              
                  person.year_of_birth BETWEEN 10 AND 19 THEN 1 END) AS age_10_to_19,
                SUM(CASE WHEN YEAR(condition.CONDITION_START_DATE) -                              
                  person.year_of_birth BETWEEN 20 AND 29 THEN 1 END) AS age_20_to_29,
                SUM(CASE WHEN YEAR(condition.CONDITION_START_DATE) -                              
                  person.year_of_birth BETWEEN 30 AND 39 THEN 1 END) AS age_30_to_39,
                SUM(CASE WHEN YEAR(condition.CONDITION_START_DATE) -                              
                  person.year_of_birth BETWEEN 40 AND 49 THEN 1 END) AS age_40_to_49,
                SUM(CASE WHEN YEAR(condition.CONDITION_START_DATE) -                              
                  person.year_of_birth BETWEEN 50 AND 59 THEN 1 END) AS age_50_to_59,
                SUM(CASE WHEN YEAR(condition.CONDITION_START_DATE) -                              
                  person.year_of_birth BETWEEN 60 AND 69 THEN 1 END) AS age_60_to_69,
                SUM(CASE WHEN YEAR(condition.CONDITION_START_DATE) -                              
                  person.year_of_birth BETWEEN 70 AND 79 THEN 1 END) AS age_70_to_79,
                SUM(CASE WHEN YEAR(condition.CONDITION_START_DATE) -                              
                  person.year_of_birth BETWEEN 80 AND 89 THEN 1 END) AS age_80_to_89,
                SUM(CASE WHEN YEAR(condition.CONDITION_START_DATE) -                              
                  person.year_of_birth >= 90 THEN 1 ELSE 0 END) AS age_90_and_more",
y = F)
disconnect(conn)
```

`r knitr::kable(t(bipolar_age), col.names = "COUNT")`


#### Racial Breakdown
```{r Bipolar Race}
conn <- connect(connectionDetails)
bipolar_race <- renderTranslateQuerySql(conn, condition_sql, ids = bipolar_concept,
                                        x = F, query = "race.RACE_CONCEPT_NAME, COUNT(DISTINCT person.PERSON_ID)", y = T, query2 = "GROUP BY race.RACE_CONCEPT_NAME ORDER BY COUNT DESC")
disconnect(conn)

bipolar_race %>%
    mutate(RACE_CONCEPT_NAME = fct_reorder(RACE_CONCEPT_NAME, COUNT, .desc = T)) %>%
    ggplot(aes(x = RACE_CONCEPT_NAME, y = COUNT)) +
    geom_bar(stat = "identity", fill = "steelblue")+
    theme_minimal() +
    labs(title = "Bipolar Race Distribution")+
    theme(axis.title.x = element_blank())
```

`r knitr::kable(bipolar_race)`

#### Gender Breakdown
```{r Bipolar Gender}
conn <- connect(connectionDetails)
bipolar_gender <- renderTranslateQuerySql(conn, condition_sql, ids = bipolar_concept,
                                        x = F, query = "gender.GENDER_CONCEPT_NAME, COUNT(DISTINCT person.PERSON_ID)", y = T, query2 = "GROUP BY gender.GENDER_CONCEPT_NAME ORDER BY COUNT DESC")
disconnect(conn)

bipolar_gender %>%
    mutate(GENDER_CONCEPT_NAME = fct_reorder(GENDER_CONCEPT_NAME, COUNT, .desc = T)) %>%
    ggplot(aes(x = GENDER_CONCEPT_NAME, y = COUNT)) +
    geom_bar(stat = "identity", fill = "steelblue")+
    theme_minimal() +
    labs(title = "Bipolar Gender Distribution")+
    theme(axis.title.x = element_blank())
```

`r knitr::kable(bipolar_gender)`


#### Care Site Breakdown
```{r Bipolar Care Site}
conn <- connect(connectionDetails)
bipolar_caresite <- renderTranslateQuerySql(conn, condition_sql, ids = bipolar_concept, x= F, query = "visit_info.VISIT_CONCEPT_NAME, COUNT(*)", y = T, query2 = "GROUP BY visit_info.VISIT_CONCEPT_NAME ORDER BY COUNT DESC")
disconnect(conn)
```

`r knitr::kable(bipolar_caresite)`

### General Trends for Depression

#### Patient Spread

This calculates the number of patients in the dataset:

```{r Depression Cohort}
conn <- connect(connectionDetails)
depression_counts <- renderTranslateQuerySql(conn, condition_sql, 
                        ids = depression_concept, x = F, y = F, 
                        query = "COUNT(*) AS RECORDS, COUNT(DISTINCT person.PERSON_ID) AS COUNT")
disconnect(conn)
```


In this dataset, there are a total of `r depression_counts$RECORDS` records for `r depression_counts$COUNT` patients.
To further examine this data, we can break them down across the following axes:

- Condition
- State
- Age
- Race
- Gender
- Care setting

#### Condition Breakdown

For the depression characterization:

1. Select the `person` table

2. Join `condition_occurrence` using `person_id`

3. Using the concept set definition for depression, filter the persons table for the persons with a depression diagnosis

4. Join the `visit_occurrence` table using `person_id` 

5. Select the demographic of interest (location, race, ethnicity, or gender)

6. Group by the demographic 

7. Return the person counts with depression for each group


```{r Depression Condition}
conn <- connect(connectionDetails)
depression_conditions <- renderTranslateQuerySql(conn, condition_sql, 
                                                 ids = depression_concept, x = F,
                                                 query = " condition.CONDITION_CONCEPT_NAME, COUNT(*) RECORD_COUNT, COUNT(DISTINCT person.PERSON_ID) PERSON_COUNT", y = T, query2 = "GROUP BY condition.CONDITION_CONCEPT_NAME ORDER BY COUNT(*) DESC")
disconnect(conn)
```


Here is the breakdown of the top depression diagnoses: `r knitr::kable(head(depression_conditions))`



#### State Breakdown
```{r Depression State}
conn <- connect(connectionDetails)
depression_state <- renderTranslateQuerySql(conn, condition_sql, ids = depression_concept, x = F, 
                        query = "location.STATE, COUNT(DISTINCT person.PERSON_ID)",
                        y = T, query2 = "GROUP BY location.STATE ORDER BY COUNT DESC")
disconnect(conn)

depression_state %>%
  mutate(STATE = fct_reorder(STATE, COUNT, .desc = F)) %>%
  top_n(20) %>%
  ggplot(aes(x = STATE, y = COUNT)) +
  geom_bar(stat = "identity", fill = "steelblue")+
  labs(title = "Depression State Distribution")+
  theme_minimal()
```



`r knitr::kable(head(depression_state, n = 10))`




#### Age Breakdown
```{r Depression Age}
conn <- connect(connectionDetails)
depression_age <- renderTranslateQuerySql(conn, 
                                          condition_sql, 
                                          ids = depression_concept, x = F, 
                                          query = "
                MIN(YEAR(condition.CONDITION_START_DATE) -                                   
                  person.year_of_birth) as min_age, 
                MAX(YEAR(condition.CONDITION_START_DATE) - person.year_of_birth) 
                  as max_age, 
                SUM(CASE WHEN YEAR(condition.CONDITION_START_DATE) - 
                  person.year_of_birth <= 9 THEN 1 END) AS age_0_9,
                SUM(CASE WHEN YEAR(condition.CONDITION_START_DATE) -                              
                  person.year_of_birth BETWEEN 10 AND 19 THEN 1 END) AS age_10_to_19,
                SUM(CASE WHEN YEAR(condition.CONDITION_START_DATE) -                              
                  person.year_of_birth BETWEEN 20 AND 29 THEN 1 END) AS age_20_to_29,
                SUM(CASE WHEN YEAR(condition.CONDITION_START_DATE) -                              
                  person.year_of_birth BETWEEN 30 AND 39 THEN 1 END) AS age_30_to_39,
                SUM(CASE WHEN YEAR(condition.CONDITION_START_DATE) -                              
                  person.year_of_birth BETWEEN 40 AND 49 THEN 1 END) AS age_40_to_49,
                SUM(CASE WHEN YEAR(condition.CONDITION_START_DATE) -                              
                  person.year_of_birth BETWEEN 50 AND 59 THEN 1 END) AS age_50_to_59,
                SUM(CASE WHEN YEAR(condition.CONDITION_START_DATE) -                              
                  person.year_of_birth BETWEEN 60 AND 69 THEN 1 END) AS age_60_to_69,
                SUM(CASE WHEN YEAR(condition.CONDITION_START_DATE) -                              
                  person.year_of_birth BETWEEN 70 AND 79 THEN 1 END) AS age_70_to_79,
                SUM(CASE WHEN YEAR(condition.CONDITION_START_DATE) -                              
                  person.year_of_birth BETWEEN 80 AND 89 THEN 1 END) AS age_80_to_89,
                SUM(CASE WHEN YEAR(condition.CONDITION_START_DATE) -                              
                  person.year_of_birth >= 90 THEN 1 ELSE 0 END) AS age_90_and_more",
                                          y = F)
disconnect(conn)
```



`r knitr::kable(t(depression_age), col.names = "COUNT")`


#### Racial Breakdown
```{r Depression Race}
conn <- connect(connectionDetails)
depression_race <- renderTranslateQuerySql(conn, condition_sql, ids = depression_concept,
                                        x = F, query = "race.RACE_CONCEPT_NAME, COUNT(DISTINCT person.PERSON_ID)", y = T, query2 = "GROUP BY race.RACE_CONCEPT_NAME ORDER BY COUNT DESC")
disconnect(conn)

depression_race %>%
    mutate(RACE_CONCEPT_NAME = fct_reorder(RACE_CONCEPT_NAME, COUNT, .desc = T)) %>%
    ggplot(aes(x = RACE_CONCEPT_NAME, y = COUNT)) +
    geom_bar(stat = "identity", fill = "steelblue")+
    theme_minimal() +
    labs(title = "Depression Race Distribution")+
    theme(axis.title.x = element_blank())
```



`r knitr::kable(depression_race)`

#### Gender Breakdown
```{r Depression Gender}
conn <- connect(connectionDetails)
depression_gender <- renderTranslateQuerySql(conn, condition_sql, ids = depression_concept,
                                        x = F, query = "gender.GENDER_CONCEPT_NAME, COUNT(DISTINCT person.PERSON_ID)", y = T, query2 = "GROUP BY gender.GENDER_CONCEPT_NAME ORDER BY COUNT DESC")
disconnect(conn)

depression_gender %>%
    mutate(GENDER_CONCEPT_NAME = fct_reorder(GENDER_CONCEPT_NAME, COUNT, .desc = T)) %>%
    ggplot(aes(x = GENDER_CONCEPT_NAME, y = COUNT)) +
    geom_bar(stat = "identity", fill = "steelblue")+
    theme_minimal() +
    labs(title = "Depression Gender Distribution")+
    theme(axis.title.x = element_blank())

```



`r knitr::kable(depression_gender)`

#### Care Site Breakdown
```{r Depression Care Site}
conn <- connect(connectionDetails)
depression_caresite <- renderTranslateQuerySql(conn, condition_sql, ids = depression_concept, x= F, query = "visit_info.VISIT_CONCEPT_NAME, COUNT(*)", y = T, query2 = "GROUP BY visit_info.VISIT_CONCEPT_NAME ORDER BY COUNT DESC")
disconnect(conn)
```



`r knitr::kable(depression_caresite)`


### General Trends for Suicidality

#### Patient Spread

This calculates the number of patients in the dataset:

```{r Suicidality Cohort}
conn <- connect(connectionDetails)
suicidality_counts <- renderTranslateQuerySql(conn, condition_sql, 
                        ids = suicidality_concept, x = F, y = F, 
                        query = "COUNT(*) AS RECORDS, COUNT(DISTINCT person.PERSON_ID) AS COUNT")
disconnect(conn)
```



In this dataset, there are a total of `r suicidality_counts$RECORDS` records for `r suicidality_counts$COUNT` patients.


To further examine this data, we can break them down across the following axes:

- Condition
- State
- Age
- Race
- Gender
- Care setting

#### Condition Breakdown

For the suicidality characterization:

1. Select the `person` table

2. Join `condition_occurrence` using `person_id`

3. Using the concept set definition for suicidality, filter the persons table for the persons with a suicidality diagnosis

4. Join the `visit_occurrence` table using `person_id` 

5. Select the demographic of interest (location, race, ethnicity, or gender)

6. Group by the demographic 

7. Return the person counts with suicidality for each group


```{r Suicidality Condition}
conn <- connect(connectionDetails)
suicidality_conditions <- renderTranslateQuerySql(conn, condition_sql, ids = suicidality_concept, x = F,
                                              query = " condition.CONDITION_CONCEPT_NAME, COUNT(*) RECORD_COUNT, COUNT(DISTINCT person.PERSON_ID) PERSON_COUNT", y = T, query2 = "GROUP BY condition.CONDITION_CONCEPT_NAME ORDER BY COUNT(*) DESC")
disconnect(conn)
```


Here is the breakdown of the top suicidality diagnoses: `r knitr::kable(head(suicidality_conditions))`


#### State Breakdown
```{r Suicidality State}
conn <- connect(connectionDetails)
suicidality_state <- renderTranslateQuerySql(conn, condition_sql, ids = suicidality_concept, x = F, 
                        query = "location.STATE, COUNT(DISTINCT person.PERSON_ID)",
                        y = T, query2 = "GROUP BY location.STATE ORDER BY COUNT DESC")
disconnect(conn)


suicidality_state %>%
  mutate(STATE = fct_reorder(STATE, COUNT, .desc = F)) %>%
  top_n(20) %>%
  ggplot(aes(x = STATE, y = COUNT)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Suicidality State Distribution") +
  theme_minimal()
```

`r knitr::kable(head(suicidality_state, n = 10))`

#### Age Breakdown
```{r Suicidality Age}
conn <- connect(connectionDetails)
suicidality_age <- renderTranslateQuerySql(conn, condition_sql, 
                                           ids = suicidality_concept, x = F,
                                           query = "
                MIN(YEAR(condition.CONDITION_START_DATE) -                                   
                  person.year_of_birth) as min_age, 
                MAX(YEAR(condition.CONDITION_START_DATE) - person.year_of_birth) 
                  as max_age, 
                SUM(CASE WHEN YEAR(condition.CONDITION_START_DATE) - 
                  person.year_of_birth <= 9 THEN 1 END) AS age_0_9,
                SUM(CASE WHEN YEAR(condition.CONDITION_START_DATE) -                              
                  person.year_of_birth BETWEEN 10 AND 19 THEN 1 END) AS age_10_to_19,
                SUM(CASE WHEN YEAR(condition.CONDITION_START_DATE) -                              
                  person.year_of_birth BETWEEN 20 AND 29 THEN 1 END) AS age_20_to_29,
                SUM(CASE WHEN YEAR(condition.CONDITION_START_DATE) -                              
                  person.year_of_birth BETWEEN 30 AND 39 THEN 1 END) AS age_30_to_39,
                SUM(CASE WHEN YEAR(condition.CONDITION_START_DATE) -                              
                  person.year_of_birth BETWEEN 40 AND 49 THEN 1 END) AS age_40_to_49,
                SUM(CASE WHEN YEAR(condition.CONDITION_START_DATE) -                              
                  person.year_of_birth BETWEEN 50 AND 59 THEN 1 END) AS age_50_to_59,
                SUM(CASE WHEN YEAR(condition.CONDITION_START_DATE) -                              
                  person.year_of_birth BETWEEN 60 AND 69 THEN 1 END) AS age_60_to_69,
                SUM(CASE WHEN YEAR(condition.CONDITION_START_DATE) -                              
                  person.year_of_birth BETWEEN 70 AND 79 THEN 1 END) AS age_70_to_79,
                SUM(CASE WHEN YEAR(condition.CONDITION_START_DATE) -                              
                  person.year_of_birth BETWEEN 80 AND 89 THEN 1 END) AS age_80_to_89,
                SUM(CASE WHEN YEAR(condition.CONDITION_START_DATE) -                              
                  person.year_of_birth >= 90 THEN 1 ELSE 0 END) AS age_90_and_more",
                                            y = F)
disconnect(conn)
```

`r knitr::kable(t(suicidality_age), col.names = "COUNT")`


#### Racial Breakdown
```{r Suicidality Race}
conn <- connect(connectionDetails)
suicidality_race <- renderTranslateQuerySql(conn, condition_sql, ids = suicidality_concept,
                                        x = F, query = "race.RACE_CONCEPT_NAME, COUNT(DISTINCT person.PERSON_ID)", y = T, query2 = "GROUP BY race.RACE_CONCEPT_NAME ORDER BY COUNT DESC")
disconnect(conn)

suicidality_race %>%
    mutate(RACE_CONCEPT_NAME = fct_reorder(RACE_CONCEPT_NAME, COUNT, .desc = T)) %>%
    ggplot(aes(x = RACE_CONCEPT_NAME, y = COUNT)) +
    geom_bar(stat = "identity", fill = "steelblue")+
    theme_minimal() +
    labs(title = "Suicidality Race Distribution")+
    theme(axis.title.x = element_blank())

```

`r knitr::kable(suicidality_race)`

#### Gender Breakdown
```{r Suicidality Gender}
conn <- connect(connectionDetails)
suicidality_gender <- renderTranslateQuerySql(conn, condition_sql, ids = suicidality_concept,
                                        x = F, query = "gender.GENDER_CONCEPT_NAME, COUNT(DISTINCT person.PERSON_ID)", y = T, query2 = "GROUP BY gender.GENDER_CONCEPT_NAME ORDER BY COUNT DESC")
disconnect(conn)


suicidality_gender %>%
    mutate(GENDER_CONCEPT_NAME = fct_reorder(GENDER_CONCEPT_NAME, COUNT, .desc = T)) %>%
    ggplot(aes(x = GENDER_CONCEPT_NAME, y = COUNT)) +
    geom_bar(stat = "identity", fill = "steelblue")+
    theme_minimal() +
    labs(title = "Suicidality Gender Distribution")+
    theme(axis.title.x = element_blank())
```

`r knitr::kable(suicidality_gender)`

#### Care Site Breakdown
```{r Suicidality Care Site}
conn <- connect(connectionDetails)
suicidality_caresite <- renderTranslateQuerySql(conn, condition_sql, ids = suicidality_concept, x= F, query = "visit_info.VISIT_CONCEPT_NAME, COUNT(*)", y = T, query2 = "GROUP BY visit_info.VISIT_CONCEPT_NAME ORDER BY COUNT DESC")
disconnect(conn)
```

`r knitr::kable(suicidality_caresite)`
